---
title: "Employment Trends Analysis in Canada: 2015-2019 vs. 2020-2024"
author: Ya-Nuo, Hsu and Kevin Johnson
fontsize: 12pt
mainfont: "Times New Roman"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)

employment_trends <- read.csv("~/Desktop/Stat 184/employment_trends.csv")

employment_trends_clean <- employment_trends %>%
  filter(REF_DATE >= "2015-01" & REF_DATE <= "2024-12",
         UOM != "Dollars") %>%
  mutate(
    REF_DATE = as.character(REF_DATE),
    Employment = replace_na(VALUE, 0), 
    Period = ifelse(REF_DATE < "2020-01", "Pre-Pandemic", "Post-Pandemic")
  ) %>%
  select(-c(DGUID, SYMBOL, TERMINATED, SCALAR_FACTOR, SCALAR_ID, STATUS, DECIMALS, Estimate, VECTOR, COORDINATE, UOM_ID)) %>%
  rename(
    Industry = North.American.Industry.Classification.System..NAICS., 
    Region = GEO,
    Date = REF_DATE
  )

write.csv(employment_trends_clean, "employment_trends_clean.csv", row.names = FALSE)

post_pandemic_employment_trends <- employment_trends_clean %>%
  filter(Date >= "2020-01" & Date <= "2024-12")

pre_pandemic_employment_trends <- employment_trends_clean %>%
  filter(Date >= "2015-01" & Date <= "2019-12")

View(employment_trends_clean)

# Pre-Pandemic Stacked Bar Chart
pre_stacked_data <- pre_pandemic_employment_trends %>%
  group_by(Industry) %>%
  summarise(Total_Employment = sum(Employment, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Employment / sum(Total_Employment))

# Post-Pandemic Stacked Bar Chart
post_stacked_data <- post_pandemic_employment_trends %>%
  group_by(Industry) %>%
  summarise(Total_Employment = sum(Employment, na.rm = TRUE)) %>%
  mutate(Proportion = Total_Employment / sum(Total_Employment))

# Set global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Introduction

Employment trends are a key indicator of economic performance and recovery, particularly during periods of disruption. In this report, we analyze how employment in Canada has evolved over two distinct periods: the pre-pandemic years (2015–2019) and the post-pandemic years (2020–2024). The goal is to identify changes in employment trends across industries and regions, using data spanning from 2015 to 2024.

## Literature Review

Employment trends have been extensively studied, especially during periods of economic recovery following major events like the COVID-19 pandemic. Studies have identified regional and industry-specific disparities in employment recovery post-pandemic. For instance, service-based industries often experience slower recovery compared to technology-driven sectors. This section will explore relevant literature on employment trends in Canada and similar economies, emphasizing regional and industrial differences.

## Methodology

We use a publicly available dataset covering monthly employment data from 2015 to 2024. The dataset includes fields for regions, industries, and employment numbers. Data were filtered to include only the years of interest (2015–2024) and were classified into two periods: Pre-Pandemic (2015–2019) and Post-Pandemic (2020–2024).

Statistical analysis and visualizations are used to identify trends and significant differences between these periods.

## Data Exploration

```{r data-exploration}
summary_stats <- employment_trends_clean %>%
  group_by(Period, Industry) %>%
  summarise(
    Avg_Employment = mean(Employment, na.rm = TRUE),
    Min_Employment = min(Employment, na.rm = TRUE),
    Max_Employment = max(Employment, na.rm = TRUE),
    SD_Employment = sd(Employment, na.rm = TRUE)
  )

kable(summary_stats, caption = "Summary Statistics for Employment Trends") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualization: Employment Trends Over Time

```{r results-visualizations}
employment_trends_group <- employment_trends_clean %>%
  mutate(
    Region_Group = case_when(
      Region %in% c("British Columbia", "Alberta", "Saskatchewan", "Manitoba") ~ "Western Canada",
      Region %in% c("Ontario", "Quebec") ~ "Central Canada",
      Region %in% c("New Brunswick", "Nova Scotia", "Prince Edward Island", "Newfoundland and Labrador") ~ "Atlantic Canada",
      TRUE ~ "Northern Territories"
    )
  )

ggplot(employment_trends_group, aes(x = Date, y = Employment, color = Industry, group = interaction(Region, Industry))) +
  geom_line() +
  facet_wrap(~Region_Group, scales = "free_y") +
  labs(
    title = "Employment Trends by Region Groups (2015–2024)",
    x = "Month-Year",
    y = "Employment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  scale_x_discrete(
    breaks = c("2015-01", "2017-01", "2019-01", "2021-01", "2023-01")
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_viridis_d()
```

# Visualization: Employment Comparison Pre- vs Post-Pandemic

```{r}
#Pre-Pandemic
ggplot(pre_stacked_data, aes(x = Industry, y = Proportion, fill = Industry)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Proportion of Employment by Industry (Pre-Pandemic)",
    subtitle = "2015–2019",
    x = "Industry",
    y = "Proportion of Employment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::percent)
#Post-Panedmic
ggplot(post_stacked_data, aes(x = Industry, y = Proportion, fill = Industry)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Proportion of Employment by Industry (Post-Pandemic)",
    subtitle = "2020–2024",
    x = "Industry",
    y = "Proportion of Employment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  scale_fill_viridis_d() +
  scale_y_continuous(labels = scales::percent)

```

# Results

# Conclusion

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
